---
title: "Final Project Report from Group 34"
author: "36-462 -- Maria Rodriguez"
output: 
  pdf_document: 
    highlight: tango
    latex_engine: xelatex
    df_print: kable
    keep_tex: yes
    fig_crop: yes
mainfont: Palatino
fontsize: 11pt
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=T, autodep=F, cache.comments=FALSE, 
                      warning=F, message=F, tidy=T, strip.white = TRUE)
```

Q1a: Is there any interesting structure present in the data? 

The data frame given to us contained 133 patients with 248 gene encodings. Given this high dimensionality it is extremely difficult to determine if there is any structure in the data. To solve this we explore several different dimension reduction methods to determine what sort of structure exists in the data. 

The first method explored is Principal Component Analysis (PCA), PCA is used as a technique to emphasize variation and highllight strong patterns in data. This also makes it easy to explore and visualize. [Source](http://setosa.io/ev/principal-component-analysis/) 

The two methods that apply PCA in R are princomp and prcomp. Princomp cannot be used on X because it uses the spectral decomposition approach, this would factorize X into a canonical form where it is represented by its eigenvalues and eigenvectors. Only diagnolizable matrices can be factored in this way, X is not a diagnolizable matrix.
[Source1](http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/)
[Source2](https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix)
[Source3](https://en.wikipedia.org/wiki/Diagonalizable_matrix)

Thus we apply prcomp to our data because it uses singular value decomposition, this examines the covariance and correlations between our observed patients and doesn't require X to be a diagnolizable matrix. But before this we explore it for a bit.

Load Data, Explore type, shape and NAs
```{r}
X <- read.table(file="trainX.txt",header=TRUE)
y <- read.table(file="trainy.txt",header=TRUE)

#Shape
dim(X)
#Checking to see if any column is fully 0, there's none
sum(colSums(X) == 0)
#No NA's
sum(apply(X, 1, function(x){any(is.na(x))}))

#These have really long output so we're not executing them 
#but it seems all column's are doubles
#summary(X)
#str(X)
head(X[,1:5])
```

Feature Selection
```{r}
#install.packages("caret", dependencies = TRUE)
library(ggplot2)
library(caret)

correlationMatrix <- cor(X)
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff=0.85)
print(highlyCorrelated)
```


PCA: With prcomp we consider the case where X it is scaled. From our summary output we see that they have many different ranges and these large differences in magnitude make scaling an appropriate step to take in this scenario. If we weren't to scale our variables the magnitude of certain variables would dominate to associations between variables in the sample.(Source)[https://stats.stackexchange.com/questions/268264/normalizing-all-the-variarbles-vs-using-scale-true-option-in-prcomp-in-r].
```{r}
pc.n = prcomp(X, center = T, scale = T)
names(pc.n)
summary(pc.n)
#PC1 explains 11%
#PC50 explains 90%
#PC65 explains 95%
#PC92 explains 99%
```

To visualize the dimensions we use factoextra to create a ggplot2 visualization of our PCA objects.
```{r}
install.packages("ggplot2")
install.packages("factoextra")
library(ggplot2)
library(factoextra)
```

The screeplot of PC object shows an elbow around 10 but that isn't enough variance explained. To determine the amount of PC's to retain I need to see the increasing gains in variance explained by each new PC. Using this it seems that the elbow of cumulative variance explained is passed around 60. So going down to 65 PC's seems to be the reasonable given it covers 95% of variance.
```{r}
fviz_eig(pc.n, ncp = 80)
vars = apply(pc.n$x, 2, var)
prop_cumsum = cumsum(vars / sum(vars))
qplot(seq_along(prop_cumsum), prop_cumsum) + xlab("PC") + 
  ylab("Cumulative Proportion of Variance Explained") + theme_minimal() +
  scale_x_continuous(breaks = seq(0, 133, by = 5)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05))
```

You can observe that there are two groups of variables that have similar meaning in the context of the data. The variables group on the right seem to have large positive loadings on dimension 2. On the group that is pointing left they seem to have high negative loadings on dimension 1.
There's also the variables clustered in the center lower right half, these are pointing away from the two extreme clusters which means they are unlike those two groups, they have low positives loadings on dimension 1 and dimension 2.
Looking at the contributions of variables it seems agreed upon in the variable loading plot that there's three different variable behaviors, those positive in dim2, negative on dim1, and all others.
```{r}
fviz_pca_biplot(pc.n)
fviz_pca_var(pc.n,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

```

For the patients themselves it seems there are two differing groups, one highly positive on dimension 2 and one highly negative on dimension 1 and these groups have high contributions to the data. Most other patients lie in the center of the two dimensions.
```{r}
fviz_pca_ind(pc.n,
             col.ind = "contrib", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

So settling on 65PC's we apply our clustering methods to see which works best on our data.
```{r}



dirs = pc.n$rotation
scrs = pc.n$x
dim(scrs)
pc.n$x[,1:65]
```


The proportion of variance explained by the principal component directions is a way of quanitfying how much structure of the data is being captured.


#XXXLater on when exploring LDA check the normality assumption of the data
